### 2025-02-14 paper/body_ecml.tex
fucking latex

>>>>

\subsection{Illustration}\label{illustration}

To better convey the intuition underlying our proposed method, we
illustrate different model outcomes in Example~\ref{exm-grad}.

\begin{example}[Prediction of Consumer Credit
Default]\protect\hypertarget{exm-grad}{}\label{exm-grad}

Suppose we are interested in predicting the likelihood that loan
applicants default on their credit. We have access to historical data on
previous loan takers comprised of a binary outcome variable
(\(y\in\{1=\text{default},2=\text{no default}\}\)) two input features:
1) the subjects' \emph{age}, which we define as immutable, and 2) the
subjects' existing level of \emph{debt}, which we define as mutable.

We have simulated this scenario using synthetic data with independent
features and Gaussian class-conditional densities in
Figure~\ref{fig-poc}. The four panels in Figure~\ref{fig-poc} show the
outcomes for different training procedures using the same model
architecture each time (a linear classifier). In each case, we show the
linear decision boundary (green) and the training data colored according
to their ground-truth label: orange points belong to the target class,
\(y^+=2\), blue points belong to the non-target class, \(y^-=1\). Stars
indicate counterfactuals in the target class generated at test time
using generic gradient descent for a fixed number of iterations.

In panel (a), we have trained our model conventionally, and we do not
impose mutability constraints at test time. The generated
counterfactuals are all valid, but not plausible: they are clearly
distinguishable from the ground-truth data. In panel (b), we have
trained our model with counterfactual training, once again not imposing
mutability constraints at test time. We observe that the counterfactuals
are clearly plausible, therefore meeting the first objective of
Definition~\ref{def-explainability}.

In panel (c), we have used conventional training again, this time
imposing the mutability constraint on \emph{age} at test time.
Counterfactuals are valid but involve some substantial reductions in
\emph{debt} for some individuals. By comparison, counterfactual paths
are shorter on average in panel (d), where we have used counterfactual
training and protected immutable features as described in
Section~\ref{sec-constraints}. The counterfactuals are also plausible
with respect to the mutable feature. Thus, we consider the model in
panel (d) as the most explainable according to
Definition~\ref{def-explainability}.

\end{example}

\begin{figure}

\centering{

\pandocbounded{\includegraphics[keepaspectratio]{body_ecml_files/mediabag/../paper/figures/poc.pdf}}

}

\caption{\label{fig-poc}Visual illustration of how counterfactual
training improves explainability. See Example~\ref{exm-grad} for
details.}

\end{figure}%

### 2025-02-14 paper/preprint/paper.tex
fucking latex

\usepackage{arxiv}
\usepackage{orcidlink}
\usepackage{amsmath}
\usepackage[T1]{fontenc}
\usepackage{placeins}
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\@ifpackageloaded{fontawesome5}{}{\usepackage{fontawesome5}}
\definecolor{quarto-callout-color}{HTML}{909090}
\definecolor{quarto-callout-note-color}{HTML}{0758E5}
\definecolor{quarto-callout-important-color}{HTML}{CC1914}
\definecolor{quarto-callout-warning-color}{HTML}{EB9113}
\definecolor{quarto-callout-tip-color}{HTML}{00A047}
\definecolor{quarto-callout-caution-color}{HTML}{FC5300}
\definecolor{quarto-callout-color-frame}{HTML}{acacac}
\definecolor{quarto-callout-note-color-frame}{HTML}{4582ec}
\definecolor{quarto-callout-important-color-frame}{HTML}{d9534f}
\definecolor{quarto-callout-warning-color-frame}{HTML}{f0ad4e}
\definecolor{quarto-callout-tip-color-frame}{HTML}{02b875}
\definecolor{quarto-callout-caution-color-frame}{HTML}{fd7e14}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\usepackage{amsthm}
\theoremstyle{definition}
\newtheorem{exercise}{Parameters}[section]
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[section]
\theoremstyle{plain}
\newtheorem{corollary}{Research Question}[section]
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\theoremstyle{definition}
\newtheorem{example}{Example}[section]
\theoremstyle{remark}
\AtBeginDocument{\renewcommand*{\proofname}{Proof}}
\newtheorem*{remark}{Remark}
\newtheorem*{solution}{Solution}
\newtheorem{refremark}{Remark}[section]
\newtheorem{refsolution}{Solution}[section]
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

@@

@@

We provide an illustrative example in Example~\ref{exm-grad}.

Suppose now that \(d_{\text{immtbl}}\) represents an individual's
\emph{age} and \(d_{\text{mtbl}}\) represents an individual's existing
level of credit card debt. Assume that these two features are
independent, the class conditional densities are Gaussian and we use no
other features to predict the risk of individuals defaulting on a
consumer loan using a linear classifier. We have simulated this scenario
using synthetic data in Figure~\ref{fig-mtblty}.

In panel (a) of Figure~\ref{fig-mtblty}, we have trained the linear
classifier using counterfactual training, treating both features as
mutable. The linear decision boundary is roughly equally sensitive to
both \emph{age} and existing levels of \emph{debt}.

Conversely, in panel (b) of Figure~\ref{fig-mtblty}, we have trained the
same classifier using counterfactual training, but this time treating
\emph{age} as an immutable feature. The result is a new decision
boundary that has tilted in favor of higher sensitivity to the mutable
feature (existing \emph{debt}) and lower sensitivity to the immutable
feature (\emph{age}).

\pandocbounded{\includegraphics[keepaspectratio]{paper_files/mediabag/../../paper/figures/app_mtblty.pdf}}

\caption{\label{fig-mtblty}Visual illustration of the effect of imposing
mutability constraints. See Example~\ref{exm-grad} for details.}

\begin{tcolorbox}[enhanced jigsaw, toprule=.15mm, colback=white, opacityback=0, coltitle=black, arc=.35mm, title=\textcolor{quarto-callout-warning-color}{\faExclamationTriangle}\hspace{0.5em}{Warning}, toptitle=1mm, colbacktitle=quarto-callout-warning-color!10!white, bottomrule=.15mm, colframe=quarto-callout-warning-color-frame, breakable, opacitybacktitle=0.6, rightrule=.15mm, left=2mm, bottomtitle=1mm, titlerule=0mm, leftrule=.75mm]

@@

\begin{tcolorbox}[enhanced jigsaw, toprule=.15mm, colback=white, opacityback=0, coltitle=black, arc=.35mm, title=\textcolor{quarto-callout-warning-color}{\faExclamationTriangle}\hspace{0.5em}{Warning}, toptitle=1mm, colbacktitle=quarto-callout-warning-color!10!white, bottomrule=.15mm, colframe=quarto-callout-warning-color-frame, breakable, opacitybacktitle=0.6, rightrule=.15mm, left=2mm, bottomtitle=1mm, titlerule=0mm, leftrule=.75mm]

@@

\begin{tcolorbox}[enhanced jigsaw, toprule=.15mm, colback=white, opacityback=0, coltitle=black, arc=.35mm, title=\textcolor{quarto-callout-note-color}{\faInfo}\hspace{0.5em}{Note}, toptitle=1mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm, colframe=quarto-callout-note-color-frame, breakable, opacitybacktitle=0.6, rightrule=.15mm, left=2mm, bottomtitle=1mm, titlerule=0mm, leftrule=.75mm]

@@

  full & 0.1 & \textit{Generic} & $-1.72 \cdot 10^{11}$ & $3.9 \cdot 10^{11}$ \\

@@

  \toprule
  \textbf{Objective} & \textbf{$\lambda_{\text{cost}} (\text{train})$} & \textbf{Generator} & \textbf{Value} & \textbf{Std} \\\midrule
  \endfirsthead
  \toprule
  \textbf{Objective} & \textbf{$\lambda_{\text{cost}} (\text{train})$} & \textbf{Generator} & \textbf{Value} & \textbf{Std} \\\midrule
  \endhead
  \bottomrule
  \multicolumn{5}{r}{Continuing table below.}\\
  \bottomrule
  \endfoot
  \endlastfoot
  full & 0 & \textit{ECCo} & $-5.32 \cdot 10^{3}$ & $1.21 \cdot 10^{4}$ \\
  full & 0 & \textit{Generic} & $-1.03 \cdot 10^{18}$ & $2.34 \cdot 10^{18}$ \\
  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-2.64}} & \color{blue}{\textbf{0.125}} \\
  full & 0 & \textit{REVISE} & -15.4 & 12.9 \\
  vanilla & 0 & \textit{ECCo} & -4.34 & 3.58 \\
  vanilla & 0 & \textit{Generic} & -4.41 & 3.48 \\
  vanilla & 0 & \textit{Omniscient} & -5.18 & 4.54 \\
  vanilla & 0 & \textit{REVISE} & -4.93 & 4.23 \\
  full & 0.001 & \textit{ECCo} & -362 & 811 \\
  full & 0.001 & \textit{Generic} & $-2.65 \cdot 10^{17}$ & $6.03 \cdot 10^{17}$ \\
  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.001}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-2.49}} & \color{blue}{\textbf{0.115}} \\
  full & 0.001 & \textit{REVISE} & -15.5 & 13 \\
  vanilla & 0.001 & \textit{ECCo} & -4.34 & 3.58 \\
  vanilla & 0.001 & \textit{Generic} & -4.41 & 3.48 \\
  vanilla & 0.001 & \textit{Omniscient} & -5.18 & 4.53 \\
  vanilla & 0.001 & \textit{REVISE} & -4.93 & 4.23 \\
  full & 0.1 & \textit{ECCo} & $-3.72 \cdot 10^{11}$ & $8.46 \cdot 10^{11}$ \\
  full & 0.1 & \textit{Generic} & $-4.49 \cdot 10^{14}$ & $1.02 \cdot 10^{15}$ \\
  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.1}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-2.5}} & \color{blue}{\textbf{0.112}} \\
  full & 0.1 & \textit{REVISE} & -14.6 & 12.2 \\
  vanilla & 0.1 & \textit{ECCo} & -4.34 & 3.58 \\
  vanilla & 0.1 & \textit{Generic} & -4.41 & 3.48 \\
  vanilla & 0.1 & \textit{Omniscient} & -5.18 & 4.54 \\
  vanilla & 0.1 & \textit{REVISE} & -4.93 & 4.24 \\\bottomrule

>>>>

\usepackage{arxiv}
\usepackage{orcidlink}
\usepackage{amsmath}
\usepackage[T1]{fontenc}
\usepackage{placeins}
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\@ifpackageloaded{fontawesome5}{}{\usepackage{fontawesome5}}
\definecolor{quarto-callout-color}{HTML}{909090}
\definecolor{quarto-callout-note-color}{HTML}{0758E5}
\definecolor{quarto-callout-important-color}{HTML}{CC1914}
\definecolor{quarto-callout-warning-color}{HTML}{EB9113}
\definecolor{quarto-callout-tip-color}{HTML}{00A047}
\definecolor{quarto-callout-caution-color}{HTML}{FC5300}
\definecolor{quarto-callout-color-frame}{HTML}{acacac}
\definecolor{quarto-callout-note-color-frame}{HTML}{4582ec}
\definecolor{quarto-callout-important-color-frame}{HTML}{d9534f}
\definecolor{quarto-callout-warning-color-frame}{HTML}{f0ad4e}
\definecolor{quarto-callout-tip-color-frame}{HTML}{02b875}
\definecolor{quarto-callout-caution-color-frame}{HTML}{fd7e14}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\usepackage{amsthm}
\theoremstyle{plain}
\newtheorem{corollary}{Research Question}[section]
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[section]
\theoremstyle{definition}
\newtheorem{example}{Example}[section]
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\theoremstyle{definition}
\newtheorem{exercise}{Parameters}[section]
\theoremstyle{remark}
\AtBeginDocument{\renewcommand*{\proofname}{Proof}}
\newtheorem*{remark}{Remark}
\newtheorem*{solution}{Solution}
\newtheorem{refremark}{Remark}[section]
\newtheorem{refsolution}{Solution}[section]
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

@@

\subsection{Illustration}\label{illustration}

To better convey the intuition underlying our proposed method, we
illustrate different model outcomes in Example~\ref{exm-grad}.

Suppose we are interested in predicting the likelihood that loan
applicants default on their credit. We have access to historical data on
previous loan takers comprised of a binary outcome variable
(\(y\in\{1=\text{default},2=\text{no default}\}\)) two input features:
1) the subjects' \emph{age}, which we define as immutable, and 2) the
subjects' existing level of \emph{debt}, which we define as mutable.

We have simulated this scenario using synthetic data with independent
features and Gaussian class-conditional densities in
Figure~\ref{fig-poc}. The four panels in Figure~\ref{fig-poc} show the
outcomes for different training procedures using the same model
architecture each time (a linear classifier). In each case, we show the
linear decision boundary (green) and the training data colored according
to their ground-truth label: orange points belong to the target class,
\(y^+=2\), blue points belong to the non-target class, \(y^-=1\). Stars
indicate counterfactuals in the target class generated at test time
using generic gradient descent for a fixed number of iterations.

In panel (a), we have trained our model conventionally, and we do not
impose mutability constraints at test time. The generated
counterfactuals are all valid, but not plausible: they are clearly
distinguishable from the ground-truth data. In panel (b), we have
trained our model with counterfactual training, once again not imposing
mutability constraints at test time. We observe that the counterfactuals
are clearly plausible, therefore meeting the first objective of
Definition~\ref{def-explainability}.

In panel (c), we have used conventional training again, this time
imposing the mutability constraint on \emph{age} at test time.
Counterfactuals are valid but involve some substantial reductions in
\emph{debt} for some individuals. By comparison, counterfactual paths
are shorter on average in panel (d), where we have used counterfactual
training and protected immutable features as described in
Section~\ref{sec-constraints}. The counterfactuals are also plausible
with respect to the mutable feature. Thus, we consider the model in
panel (d) as the most explainable according to
Definition~\ref{def-explainability}.

\pandocbounded{\includegraphics[keepaspectratio]{paper_files/mediabag/../../paper/figures/poc.pdf}}

\caption{\label{fig-poc}Visual illustration of how counterfactual
training improves explainability. See Example~\ref{exm-grad} for
details.}

@@

\begin{tcolorbox}[enhanced jigsaw, arc=.35mm, bottomtitle=1mm, coltitle=black, opacitybacktitle=0.6, rightrule=.15mm, toprule=.15mm, toptitle=1mm, colbacktitle=quarto-callout-warning-color!10!white, colback=white, breakable, left=2mm, leftrule=.75mm, titlerule=0mm, title=\textcolor{quarto-callout-warning-color}{\faExclamationTriangle}\hspace{0.5em}{Warning}, bottomrule=.15mm, opacityback=0, colframe=quarto-callout-warning-color-frame]

@@

\begin{tcolorbox}[enhanced jigsaw, arc=.35mm, bottomtitle=1mm, coltitle=black, opacitybacktitle=0.6, rightrule=.15mm, toprule=.15mm, toptitle=1mm, colbacktitle=quarto-callout-warning-color!10!white, colback=white, breakable, left=2mm, leftrule=.75mm, titlerule=0mm, title=\textcolor{quarto-callout-warning-color}{\faExclamationTriangle}\hspace{0.5em}{Warning}, bottomrule=.15mm, opacityback=0, colframe=quarto-callout-warning-color-frame]

@@

\begin{tcolorbox}[enhanced jigsaw, arc=.35mm, bottomtitle=1mm, coltitle=black, opacitybacktitle=0.6, rightrule=.15mm, toprule=.15mm, toptitle=1mm, colbacktitle=quarto-callout-note-color!10!white, colback=white, breakable, left=2mm, leftrule=.75mm, titlerule=0mm, title=\textcolor{quarto-callout-note-color}{\faInfo}\hspace{0.5em}{Note}, bottomrule=.15mm, opacityback=0, colframe=quarto-callout-note-color-frame]

@@

  full & 0.1 & \textit{Generic} & $-1.71 \cdot 10^{11}$ & $3.9 \cdot 10^{11}$ \\

@@

  \toprule
  \textbf{Objective} & \textbf{$\lambda_{\text{cost}} (\text{train})$} & \textbf{Generator} & \textbf{Value} & \textbf{Std} \\\midrule
  \endfirsthead
  \toprule
  \textbf{Objective} & \textbf{$\lambda_{\text{cost}} (\text{train})$} & \textbf{Generator} & \textbf{Value} & \textbf{Std} \\\midrule
  \endhead
  \bottomrule
  \multicolumn{5}{r}{Continuing table below.}\\
  \bottomrule
  \endfoot
  \endlastfoot
  full & 0 & \textit{ECCo} & $-5.32 \cdot 10^{3}$ & $1.21 \cdot 10^{4}$ \\
  full & 0 & \textit{Generic} & $-1.03 \cdot 10^{18}$ & $2.34 \cdot 10^{18}$ \\
  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-2.64}} & \color{blue}{\textbf{0.125}} \\
  full & 0 & \textit{REVISE} & -15.4 & 12.9 \\
  vanilla & 0 & \textit{ECCo} & -4.34 & 3.58 \\
  vanilla & 0 & \textit{Generic} & -4.41 & 3.48 \\
  vanilla & 0 & \textit{Omniscient} & -5.18 & 4.54 \\
  vanilla & 0 & \textit{REVISE} & -4.93 & 4.23 \\
  full & 0.001 & \textit{ECCo} & -362 & 811 \\
  full & 0.001 & \textit{Generic} & $-2.65 \cdot 10^{17}$ & $6.04 \cdot 10^{17}$ \\
  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.001}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-2.49}} & \color{blue}{\textbf{0.115}} \\
  full & 0.001 & \textit{REVISE} & -15.5 & 13 \\
  vanilla & 0.001 & \textit{ECCo} & -4.34 & 3.58 \\
  vanilla & 0.001 & \textit{Generic} & -4.41 & 3.48 \\
  vanilla & 0.001 & \textit{Omniscient} & -5.18 & 4.53 \\
  vanilla & 0.001 & \textit{REVISE} & -4.93 & 4.23 \\
  full & 0.1 & \textit{ECCo} & $-3.72 \cdot 10^{11}$ & $8.46 \cdot 10^{11}$ \\
  full & 0.1 & \textit{Generic} & $-4.48 \cdot 10^{14}$ & $1.02 \cdot 10^{15}$ \\
  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.1}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-2.5}} & \color{blue}{\textbf{0.112}} \\
  full & 0.1 & \textit{REVISE} & -14.6 & 12.2 \\
  vanilla & 0.1 & \textit{ECCo} & -4.34 & 3.58 \\
  vanilla & 0.1 & \textit{Generic} & -4.41 & 3.48 \\
  vanilla & 0.1 & \textit{Omniscient} & -5.18 & 4.54 \\
  vanilla & 0.1 & \textit{REVISE} & -4.93 & 4.24 \\\bottomrule

### 2025-02-14 paper/sections/method.qmd
fucking latex

>>>>

## Illustration

To better convey the intuition underlying our proposed method, we illustrate different model outcomes in @exm-grad.

::: {#exm-grad}

## Prediction of Consumer Credit Default

Suppose we are interested in predicting the likelihood that loan applicants default on their credit. We have access to historical data on previous loan takers comprised of a binary outcome variable ($y\in\{1=\text{default},2=\text{no default}\}$) two input features: 1) the subjects' *age*, which we define as immutable, and 2) the subjects' existing level of *debt*, which we define as mutable.

We have simulated this scenario using synthetic data with independent features and Gaussian class-conditional densities in @fig-poc. The four panels in @fig-poc show the outcomes for different training procedures using the same model architecture each time (a linear classifier). In each case, we show the linear decision boundary (green) and the training data colored according to their ground-truth label: orange points belong to the target class, $y^+=2$, blue points belong to the non-target class, $y^-=1$. Stars indicate counterfactuals in the target class generated at test time using generic gradient descent for a fixed number of iterations.

In panel (a), we have trained our model conventionally, and we do not impose mutability constraints at test time. The generated counterfactuals are all valid, but not plausible: they are clearly distinguishable from the ground-truth data. In panel (b), we have trained our model with counterfactual training, once again not imposing mutability constraints at test time. We observe that the counterfactuals are clearly plausible, therefore meeting the first objective of @def-explainability.

In panel (c), we have used conventional training again, this time imposing the mutability constraint on *age* at test time. Counterfactuals are valid but involve some substantial reductions in *debt* for some individuals. By comparison, counterfactual paths are shorter on average in panel (d), where we have used counterfactual training and protected immutable features as described in @sec-constraints. The counterfactuals are also plausible with respect to the mutable feature. Thus, we consider the model in panel (d) as the most explainable according to @def-explainability.

![Visual illustration of how counterfactual training improves explainability. See @exm-grad for details.](/paper/figures/poc.svg){#fig-poc}

### 2025-02-14 paper/sections/other/constraints.qmd
fucking latex

We provide an illustrative example in @exm-grad.

::: {#exm-grad}

## Prediction of Consumer Credit Default

Suppose now that $d_{\text{immtbl}}$ represents an individual's *age* and $d_{\text{mtbl}}$ represents an individual's existing level of credit card debt. Assume that these two features are independent, the class conditional densities are Gaussian and we use no other features to predict the risk of individuals defaulting on a consumer loan using a linear classifier. We have simulated this scenario using synthetic data in @fig-mtblty.

In panel (a) of @fig-mtblty, we have trained the linear classifier using counterfactual training, treating both features as mutable. The linear decision boundary is roughly equally sensitive to both *age* and existing levels of *debt*.

Conversely, in panel (b) of @fig-mtblty, we have trained the same classifier using counterfactual training, but this time treating *age* as an immutable feature. The result is a new decision boundary that has tilted in favor of higher sensitivity to the mutable feature (existing *debt*) and lower sensitivity to the immutable feature (*age*).

![Visual illustration of the effect of imposing mutability constraints. See @exm-grad for details.](/paper/figures/app_mtblty.svg){#fig-mtblty}

@@

function plot_db(model, x; _xlab="Existing Debt", _ylab="Age")

@@

    # Data and decision boundary:
    coeff = get_log_reg_params(model)
    db = get_decision_boundary(coeff)
    plt = Plots.scatter(
        ce_data, 
        title="Iteration $(_itercount)",
        # label=["1=Default" "2=No Default"], 
        # legend_position=:topright, 
        bottom_margin = 5mm, 
        left_margin = 5mm, 
        xlab=xlab, ylab=ylab,  
        axis=nothing, 
        legend=false
    )
    Plots.abline!(plt, db.slope, db.intercept; lw=5, label="Dec. Boundary")

    # Gradient field:
    targets = Flux.onehotbatch(fill(2,size(x,2)), [1,2])
    samples = ce_data.X[:,ce_data.output_encoder.labels.==2][:,rand(1:end,size(x,2))]
    u = []
    v = []
    η = 0.1
    for (i,ce) in enumerate(eachcol(x))
        grads = Flux.gradient((m,x,s,t) -> implausibility(m,x,s,t)[1], model, ce, samples[:,i], targets[:,i])
        g = grads[1].layers[1].weight
        push!(u, η * g[1])
        push!(v, η * g[2])
    end
    Plots.quiver!(x1, x2, quiver=(u, v))

    # Counterfactuals:
    if !any(isnothing.(x))
        yhat = [argmax(y) for y in eachcol(model(x))]
        if any(yhat.==2)
            Plots.scatter!(x[1,yhat.==2], x[2,yhat.==2], label=["CE (y⁺=1)" "CE (y⁺=2)"], ms=15, shape=:star, color=yhat[yhat.==2], group=yhat[yhat.==2])
        end
    end
    display(plt)

    if _itercount == 5
        push!(_plts, plt)
    end
    global _itercount += 1
end

# Data:
specs = [
    ("(a)", VanillaObjective(needs_ce=true), ["both", "both"]),
    ("(b)", EnergyDifferentialObjective(lambda=[1.0,1.0,0.0]), ["both", "both"]),
    ("(c)", EnergyDifferentialObjective(lambda=[1.0,1.0,0.0]), ["both", "none"]),
]
```

@@

    # Globals for the callback:
    global _constraint = constraint
    global _plts = []
    global _itercount = 1

@@

    model, logs = counterfactual_training(
        obj,
        model,
        generator,
        train_set,
        opt_state;
        val_set = val_set,
        nepochs = 5,
        mutability = Symbol.(constraint),
        callback = plot_db,
        nce=50,
        convergence=conv, 
    )

    push!(models, model)
    push!(_plts, current())
    plt = Plots.plot(_plts..., plot_title = title, layout=(2,1))
    push!(plts, plt)
end
```

>>>>

@@

function plot_db(model, ces; _xlab="Debt", _ylab="Age")

    x = (ce -> ce.counterfactual).(ces) |> xs -> reduce(hcat, xs)
    x0 = (ce -> ce.factual).(ces) |> xs -> reduce(hcat, xs)

@@

    # Data and decision boundary:
    coeff = get_log_reg_params(model)
    db = get_decision_boundary(coeff)
    plt = Plots.scatter(
        ce_data, 
        # label=["1=Default" "2=No Default"], 
        # legend_position=:topright, 
        xlab=xlab, ylab=ylab,  
        axis=nothing, 
        legend=false,
        title=_title
    )
    Plots.abline!(plt, db.slope, db.intercept; lw=5, label="Dec. Boundary")

    # Counterfactuals:
    if !any(isnothing.(x))
        yhat = [argmax(y) for y in eachcol(model(x))]
        yhat0 = [argmax(y) for y in eachcol(model(x0))]
        if any(yhat.==2)

            # Paths:
            u = []
            v = []
            for (i,ce) in enumerate(eachcol(x))
                Δ = ce - x0[:,i]
                push!(u, Δ[1])
                push!(v, Δ[2])
            end
            Plots.quiver!(x0[1,yhat.==2], x0[2,yhat.==2], quiver=(u[yhat.==2], v[yhat.==2]), color=1)

            # End points:
            Plots.scatter!(x[1,yhat.==2], x[2,yhat.==2], label=["CE (y⁺=1)" "CE (y⁺=2)"], ms=15, shape=:star, color=yhat[yhat.==2], group=yhat[yhat.==2], mscolor=yhat0[yhat.==2])

        end
    end
    display(plt)
end

# Data:
specs = [
    ("(a)", VanillaObjective(needs_ce=true), ["both", "both"]),
    ("(b)", EnergyDifferentialObjective(lambda=[1.0,1.0,0.0]), ["both", "both"]),
    ("(c)", VanillaObjective(needs_ce=true), ["both", "none"]),
    ("(d)", EnergyDifferentialObjective(lambda=[1.0,1.0,0.0]), ["both", "none"]),
]
```

@@

    # Globals for the callback:
    global _title = title
    global _constraint = constraint

@@

    model, logs = counterfactual_training(
        obj,
        model,
        generator,
        train_set,
        opt_state;
        val_set = val_set,
        nepochs = 50,
        mutability = Symbol.(constraint),
        callback = plot_db,
        nce=50,
        convergence=conv, 
    )

    push!(models, model)
    push!(plts, current())
end
```

plt = Plots.plot(
    plts..., 
    layout=(1,4), 
    size=(1150,250),  
    left_margin = 10mm, 
    bottom_margin = 5mm,
    top_margin = 3mm,
    right_margin = 10mm,
)
display(plt)
Plots.savefig(plt, "paper/figures/poc.svg")
```
